---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
```{r activity}
library(dplyr)
library(ggplot2)
library(plyr)
library(lattice)

activity <- read.csv("activity.csv")
activity <- transform(activity, date = as.Date(date))
```


## What is mean total number of steps taken per day?

```{r}
activity$day <- weekdays(activity$date)
TotSteps_by_day <- aggregate(steps~date, activity, sum)
```

Histogram of the total number of steps taken each day:

```{r}
hist(TotSteps_by_day$steps, xlab = "Number of Steps", ylab = "Number of days", main = "Total number of steps taken per Day" , labels = TRUE, ylim = c(0,35))
```

Mean of the total number of steps taken per day:

```{r}
meanSteps <- as.integer(mean(TotSteps_by_day$steps))
meanSteps

```

Median of the total number of steps taken per day:

```{r}
medSteps <- as.integer(median(TotSteps_by_day$steps))
medSteps
```


## What is the average daily activity pattern?
```{r}
avgSteps_perInt <- aggregate(steps~interval, activity, mean)
head(avgSteps_perInt)
```

Time-series Plot:

```{r}
with(avgSteps_perInt, plot(interval, steps, type= "l", xlab= "Interval", ylab = "Avg. Number of Steps", ylim = c(0,250)))
```

5-minute interval with the maximum number of steps:

```{r}
avgSteps_perInt[which.max(avgSteps_perInt[,2]), 1]
```



## Imputing missing values

Total number of missing values in the dataset:

```{r}
naTotal <- sum(is.na(activity))
naTotal
```

Devise a strategy for filling in all of the missing values in the dataset:

```{r}
avgSteps_dayInt <- ddply(activity, .(interval, day), summarize, Avg= mean(steps, na.rm = TRUE))

naData <- activity[is.na(activity$steps),]
newData <- merge(naData, avgSteps_dayInt, by = c("interval", "day"))

newData2 <- newData[, c(5, 4, 1, 2)]
colnames(newData2) <- c("steps", "date", "interval", "day")

activity1 <- rbind(activity[!is.na(activity$steps),], newData2)
```

Histogram of the total number of steps taken each day:

```{r}
TotSteps_day1 <- aggregate(steps~date, activity1, sum)
hist(TotSteps_day1$steps, xlab="Number of Steps per day", ylab="Number of Days", main="Number of Steps taken each day after NAs imputed", labels=TRUE, ylim = c(0,42))
```


Mean of the total number of steps taken per day:

```{r}
mean_imputed <- as.integer(mean(TotSteps_day1$steps))
mean_imputed
```

Median of the total number of steps taken per day:

```{r}
median_imputed <- as.integer(median(TotSteps_day1$steps))
median_imputed
```

The new mean of the imputed data is 10821 steps compared to the old mean of 10766 steps. That creates a difference of 55 steps on average per day.

The new median of the imputed data is 11015 steps compared to the old median of 10765 steps. That creates a difference of 250 steps for the median.
However, the overall shape of the distribution has not changed.


## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday” and “weekend”: 

```{r}
activity2 <- activity1 %>% mutate(dayType = ifelse(weekdays(activity1$date)== "Saturday" | weekdays(activity1$date) == "Sunday", "Weekend", "Weekday"))
```

Panel plot containing a time series plot:

```{r}
avgSteps_DayTypeIn <- ddply(activity2, .(interval, dayType), summarize, Avg = mean(steps))


with(avgSteps_DayTypeIn, xyplot(Avg~interval|dayType, type="l", layout = c(1,2), main = "Average Steps per Interval based on Type of Day", xla ="Interval", ylab = "Avg. Number of Steps"))

```

Yes, the step activity trends are different based on whether the day occurs on a weekend or not. This may be due to people having an increased opportunity for activity beyond normal work hours for those who work during the week.